---
title: "Homework 5"
format: html
editor: visual
---


## Task 1: Conceptual Questions

1. What is the purpose of using cross-validation when fitting a random forest model?
Cross-validation helps assess the performance of a random forest model by iteratively splitting the data into training and validation sets, ensuring robustness in model evaluation and hyperparameter tuning.

2. Describe the bagged tree algorithm.
Bagging (Bootstrap Aggregating) involves creating multiple bootstrap samples from the original dataset, training a decision tree on each sample, and then aggregating their predictions (e.g., averaging for regression or voting for classification) to reduce variance and improve model accuracy.

3. What is meant by a general linear model?
A general linear model (GLM) is a statistical model that assumes a linear relationship between the dependent variable and predictor variables, with normally distributed errors.

4. When fitting a multiple linear regression model, what does adding an interaction term do? That is,
what does it allow the model to do differently as compared to when it is not included in the model?
Adding an interaction term allows the model to capture non-additive effects between predictors, meaning the relationship between the dependent variable and one predictor variable may vary depending on the value of another predictor.

5. Why do we split our data into a training and test set?
Splitting data into training and test sets helps evaluate the performance of a predictive model on unseen data, thereby estimating how well the model generalizes to new, unseen observations and avoiding overfitting to the training data.


## Task 2
```{r}
library(readxl)
library(dplyr)
library(ggplot2)
library(caret)
library(GGally)
library(tidyr)
```


## 1. Quickly understand your data. Check on missingness and summarize the data, especially with respect
to the relationships of the variables to HeartDisease.
```{r}
#Read in data
heart <- read_excel("heart.xls")

#Understand the structure of the dataset
str(heart)
summary(heart)

#Check for missing values
missing_values <- sapply(heart, function(x) sum(is.na(x)))
print(missing_values)

# Summary statistics for numeric variables
numeric_summary <- heart %>% 
  select_if(is.numeric) %>%
  summary()
print(numeric_summary)

# Count of HeartDisease cases
heart_disease_counts <- heart %>% 
  group_by(HeartDisease) %>% 
  summarise(count = n())
print(heart_disease_counts)

# Correlation matrix for numeric predictors
numeric_vars <- heart %>% select_if(is.numeric)
cor_matrix <- cor(numeric_vars, use = "complete.obs")
print(cor_matrix)
# Strong correlations with HeartDisease indicate potential significance so age, restingbp, fastingbs and oldpeak are strongest indicators

```



## 2. Create a new variable that is a factor version of the HeartDisease variable (if needed, this depends on
how you read in your data). Remove the ST_Slope variable and the original HeartDisease variable
(if applicable).
```{r}
#Create a factor version of the HeartDisease variable
heart <- heart %>%
  mutate(HeartDiseaseFactor = as.factor(HeartDisease))

#Remove the ST_Slope variable and the original HeartDisease variable
heart <- heart %>%
  select(-ST_Slope, -HeartDisease)

```


## 3. We’ll be doing a kNN model below to predict whether or not someone has heart disease. To use
kNN we generally want to have all numeric predictors (although we could try to create our own loss
function as an alternative). In this case we have some categorical predictors still in our data set: Sex,
ExerciseAngina ChestPainType, and RestingECG. 

Create dummy columns corresponding to the values of these three variables for use in our kNN fit. The
caret vignette has a function to help us out here. You should use dummyVars() and predict() to create
new columns. Then add these columns to our data frame.

```{r}
# Create dummy variables for categorical predictors
dummies <- dummyVars("~ Sex + ExerciseAngina + ChestPainType + RestingECG", data = heart)
dummy <- predict(dummies, newdata = heart)

# Convert the dummy data to a data frame
dummy <- as.data.frame(dummy)

# Add the dummy variables to the dataset
heart <- cbind(heart, dummy)

# Remove the original categorical predictors
heart <- heart %>%
  select(-Sex, -ExerciseAngina, -ChestPainType, -RestingECG)


```


## Split your Data
```{r}
# Split the data into training and testing sets
set.seed(123)  # Setting seed for reproducibility
trainIndex <- createDataPartition(heart$HeartDiseaseFactor, p = 0.8, 
                                  list = FALSE, 
                                  times = 1)
heart_train <- heart[trainIndex,]
heart_test  <- heart[-trainIndex,]

# Verify the split
cat("Training set dimensions:", dim(heart_train), "\n")
cat("Test set dimensions:", dim(heart_test), "\n")
```


## KNN Model
```{r}
# Set up the train control for cross-validation
train_control <- trainControl(method = "repeatedcv", 
                              number = 10, 
                              repeats = 3)

# Set up the tuning grid for k values
tune_grid <- expand.grid(k = 1:40)

# Step 9: Train the kNN model
knn_model <- train(HeartDiseaseFactor ~ ., 
                   data = heart_train, 
                   method = "knn", 
                   trControl = train_control, 
                   tuneGrid = tune_grid, 
                   preProcess = c("center", "scale"))  #preprocessing data here

# Print model summary
print(knn_model)

# Make predictions
knn_predictions <- predict(knn_model, newdata = heart_test)

# Step 11: Evaluate the model
conf_matrix <- confusionMatrix(knn_predictions, heart_test$HeartDiseaseFactor)
print(conf_matrix)
```



## Logistic Regression
```{r}
# Define the models
# Define logistic regression models
model1_formula <- HeartDiseaseFactor ~ Age + RestingBP + Cholesterol + MaxHR
model2_formula <- HeartDiseaseFactor ~ Age + RestingBP + Cholesterol + MaxHR + factor(SexM) + factor(ExerciseAnginaN) + factor(ChestPainTypeASY) + factor(RestingECGNormal)
model3_formula <- HeartDiseaseFactor ~ Age + RestingBP + Cholesterol + MaxHR + I(Age^2) + RestingBP:Cholesterol

model1_train <- train(
  model1_formula,
  heart_train, 
  method = "knn",
  trControl = train_control,
  tuneGrid = tune_grid,
  preProcess = c("center", "scale")
)

model2_train <- train(
  model2_formula,
  heart_train, 
  method = "knn",
  trControl = train_control,
  tuneGrid = tune_grid,
  preProcess = c("center", "scale")
)

model3_train <- train(
  model3_formula,
  heart_train, 
  method = "knn",
  trControl = train_control,
  tuneGrid = tune_grid,
  preProcess = c("center", "scale")
)

# Extract resamples into a list
results <- resamples(list(model1 = model1_train, model2 = model2_train, model3 = model3_train))
print(results)

# Summarize the results
results <- summary(results)
print(results)

# Extract accuracies
model_accuracies <- results$statistics$Accuracy[, "Mean"]

# Identify the best model
best_model_name <- names(which.max(model_accuracies))
print(paste("Best model:", best_model_name))

# Retrieve the best model
best_model_train <- get(paste0(best_model_name, "_train"))
print(best_model_train)

# Evaluate the best model on the test set
predictions <- predict(best_model_train, newdata = heart_test)
conf_matrix <- confusionMatrix(predictions, heart_test$HeartDiseaseFactor)
print(conf_matrix)
```

## Tree Models

## Decision Tree Model
```{r}
library(caret)

# Define the formula for the decision tree model
tree_formula <- HeartDiseaseFactor ~ Age + RestingBP + Cholesterol + MaxHR + factor(SexM) + factor(ExerciseAnginaN) + factor(ChestPainTypeASY) + factor(RestingECGNormal)

# Define tuning grid for rpart (cp values)
cp_values <- seq(0, 0.1, by = 0.001)
tune_grid2 <- expand.grid(cp = cp_values)

# Train the decision tree model
set.seed(1234)
model_rpart_train <- train(
  tree_formula,
  data = heart_train,
  method = "rpart",
  trControl = train_control,
  tuneGrid = tune_grid2,
  preProcess = c("center", "scale")
)

# Print the best model and its parameters
print(model_rpart_train)

# Evaluate the decision tree model on the test set
predictions_rpart <- predict(model_rpart_train, newdata = heart_test)
conf_matrix_rpart <- confusionMatrix(predictions_rpart, heart_test$HeartDiseaseFactor)
print(conf_matrix_rpart)

```
##Rain Forest Model
```{r}

# Set up empty var for tuning.
num_predictors <- length(all.vars(tree_formula)) - 1  

# New Tuning for this tree fit
tuneGrid <- expand.grid(mtry = 1:num_predictors)

set.seed(1234)
model_rf_train <- train(
  tree_formula,
  data = heart_train,
  method = "rf",
  trControl = train_control,
  tuneGrid = tuneGrid,
  preProcess = c("center", "scale")
)

# Print the best model and its parameters
print(model_rf_train)

# Evaluate the decision tree model on the test set
predictions_rpart <- predict(model_rf_train, newdata = heart_test)
conf_matrix_rpart <- confusionMatrix(predictions_rpart, heart_test$HeartDiseaseFactor)
print(conf_matrix_rpart)


```


##Booted Tree Model

```{r}
# Define the tuning grid for the boosted tree
tuneGrid2 <- expand.grid(
  n.trees = c(25, 50, 100, 200),
  interaction.depth = c(1, 2, 3),
  shrinkage = 0.1,
  n.minobsinnode = 10
)


# Train the boosted tree model using caret
set.seed(1234)
model_gbm_train <- train(
  tree_formula,
  data = heart_train,
  method = "gbm",
  trControl = train_control,
  tuneGrid = tuneGrid2,
  preProcess = c("center", "scale"),
  verbose = FALSE
)

# Print the best model and its parameters
print(model_gbm_train)

```


##Evaluate
```{r}
# Evaluate on test set and print confusion matrices
# Decision Tree (rpart)
predictions_rpart <- predict(model_rpart_train, newdata = heart_test)
conf_matrix_rpart <- confusionMatrix(predictions_rpart, heart_test$HeartDiseaseFactor)
print("Confusion Matrix for Decision Tree (rpart):")
print(conf_matrix_rpart)

# Random Forest (rf)
predictions_rf <- predict(model_rf_train, newdata = heart_test)
conf_matrix_rf <- confusionMatrix(predictions_rf, heart_test$HeartDiseaseFactor)
print("Confusion Matrix for Random Forest (rf):")
print(conf_matrix_rf)

# Boosted Tree (gbm)
predictions_gbm <- predict(model_gbm_train, newdata = heart_test)
conf_matrix_gbm <- confusionMatrix(predictions_gbm, heart_test$HeartDiseaseFactor)
print("Confusion Matrix for Boosted Tree (gbm):")
print(conf_matrix_gbm)
```


##Wrap Up
The model with the highest accuracy value indicates which model's predictions aligned the closest with the actual outcomes in the test data. Amongst the three models, the one with the highest accuracy was model 2. 
